Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-09-15 iso-int/rc-iso-int.tex
several corrections

\item Operations {{fold}} and {{unfold}} are specific to each recursive
  data type and are drawn as triangular boxes. For instance,
  Sec. \ref{sec:bounded} shows {{fold}}/{{unfold}} isomorphisms for
  numbers, {{nat=@@x.(1+x)}}.

@@

\paragraph*{Universality.} 
As an example, consider the implementation of the Toffoli gate below,
which establishes that {{langRevT}} is complete for combinational
circuits. This encoding uses the type {{1+1}} as {{bool}} and values
{{left ()}} and {{right()}} as {{true}} and {{false}}.  Boolean
negation not is simply {{swap+}}. The Toffoli gate takes three boolean
inputs: if the first two inputs are {{true}} then the third is
negated.

@@

@@

Our goal is to implement this abstract machine in {{langRevT}}. We start by
writing a combinator {{c : nat * nat <-> nat * nat}} that executes one step
of the machine transition. Here {{nat}} is an abbreviation for the recursive
type {{@@x.1+x}}. The combinator, when iterated, should map {{(n,0)}} to
{{(0,n)}} by precisely mimicking the steps of the abstract machine. Let us
analyze this desired combinator {{c}} in detail starting from its interface:

@@

The triangle on the left denotes the {{unfold}} isomorphism, which works as
follows: If the number {{n}} is zero, the output is the top branch which has
type~{{1}}. If the number is non-zero, the output is on the bottom branch
and has value {{n-1}}. The triangle on the right is the dual {{fold}}
isomorphism.  We now use distribution to isolate each possible machine state:

@@

\item \emph{Diverge.}  Another standard approach in dealing with stuck states
  is to make the machine diverge or leave the output undefined or
  unobservable in some way. If the specific case of the machine above, we can
  use a primitive {{add1}} whose dual {{sub1}} is undefined when applied to
  {{0}}. (Such a primitive is definable in
  {{langRevT}}~\cite[Sec.~4.2]{James:2012:IE:2103656.2103667}.)

>>>>

\item Operations {{fold}} and {{unfold}} are specific to each recursive
  data type and are drawn as triangular boxes. For instance,
  Sec. \ref{sec:numeric-ops} shows {{fold}}/{{unfold}} isomorphisms for
  numbers, {{nat=@@x.(1+x)}}.

@@

\subsection{Universality.} 
As an example, consider the implementation of the Toffoli gate below,
which establishes that {{langRevT}} is complete for combinational
circuits. This encoding uses the type {{1+1}} as {{bool}} and values
{{left ()}} and {{right()}} as {{true}} and {{false}}.  Boolean
negation not is simply {{swap+}}. The Toffoli gate takes three boolean
inputs: if the first two inputs are {{true}} then the third is
negated.

@@

\subsection{Numeric Operations.}
\label{sec:numeric-ops}

In {{langRevT}} we encode numbers, {{nat = 0 ~|~ n + 1}}, using the
recursive type {{@@x.1+x}}. The {{fold}}/{{unfold}} isomorphisms for
{{nat}} are {{unfold : nat <-> 1+nat : fold}}.
\begin{multicols}{3}
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/unfold_nat.pdf}
}
\end{center}

\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/fold_nat.pdf}
}
\end{center}

\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/add1.pdf}
}
\end{center}
\end{multicols}
The triangle on the left denotes the {{unfold}} isomorphism, which
works as follows: If the number {{n}} is zero, the output is the top
branch which has type~{{1}}. If the number is non-zero, the output is
on the bottom branch and has value {{n-1}}. The triangle in the middle
is the dual {{fold}} isomorphism.

As explained in previous work
\cite[Sec.~4.2]{James:2012:IE:2103656.2103667}, numeric addition and
subtraction can be expressed in {{langRevT}} as partial
isomorphisms. The combinator on the right above is
{{add1:nat<->nat}}. It acts like the successor function returning
{{n+1}} for all inputs {{n}}. Its adjoint {{sub1:nat<->nat}}, which we
get by flipping the diagram right to left, diverges on input {{0}} and
returns {{n-1}} for all other inputs {{n}}.

@@

Our goal is to implement this abstract machine in {{langRevT}}. We start by
writing a combinator {{c : nat * nat <-> nat * nat}} that executes one step
of the machine transition. 
%% Here {{nat}} is an abbreviation for the recursive
%% type {{@@x.1+x}}. 
The combinator, when iterated, should map {{(n,0)}} to
{{(0,n)}} by precisely mimicking the steps of the abstract machine. Let us
analyze this desired combinator {{c}} in detail starting from its interface:

@@

We now use distribution to isolate each possible machine state:

@@

\item \emph{Diverge.}  Another standard approach in dealing with stuck states
  is to make the machine diverge or leave the output undefined or
  unobservable in some way. If the specific case of the machine above, we can
  use the primitive {{add1}} whose dual {{sub1}} is undefined when applied to
  {{0}} (see Sec.~\ref{sec:numeric-ops}).
